A Short and Mechanized Consistency Proof for a Dependent Type Theory

### 2023-12-18 paper/paper.tex
Add intro

\section{Introduction}
NbE in Coq~\citep{nbeincoq}

>>>>

\section{Introduction}
% Depending on its application, we care about certain metatheoretic
% properties about a type system. As a programming language, we may care
% about type soundness, which states that a well-typed never gets stuck
% during evaluation.
When a type system is used as a program logic where terms encode proofs, we
want our type system to be logically consistent, meaning
that the empty type is not inhabited.
This paper is specifically
about establishing logical consistency for a fully dependently typed
system with an infinite universe hierarchy and support for large
elimination. % The type system, presented in Section~\ref{sec:spec}, is
% most similar to Martin-Lof's predicate type theory with the minor
% difference that type conversion is based on untyped equality.

The proof of various dependently typed systems, including Martin-Lof's
type theory, has long been
established in the literature and therefore the consistency result of
this paper is not novel. Rather,
our goal is to present the proof in a form that is digestable by a
working type theorist and can be more readily mechanized in a proof
assistant. Compared to existing efforts at mechanizing logical
consistency or stronger properties such as existence of normal
form~\citep{nbeincoq},
decidable type checking~\citep{decagda}, our work is minimal since it requires very
little scaffolding and therefore results in an extremely succinct
proof of under 1000 lines of manually written Coq code for a dependent
type theory that is reasonably complete in its features.

The key technique that underlies our consistency proof is proof by
logical relation. In Section~\ref{sec:logrelstlc}, we review the
technique in the context of the simply typed lambda calculus. In
Section~\ref{sec:spec}, we present the dependent type theory of
interest. In Section~\ref{sec:logreldep}, we give the definition
of the logical relation for the dependent type theory. Rather than
presenting the logical relation as an inductive-recursive definition,
we use the more elementary concept of a partial function to capture
the interpretation of types. The alternative representation requires us
to show that the set of equations indeed defines a partial function;
that is, for each input, there should always be a unique
output.
From the interpretation function, we can define the semantic
typing judgment for the set of lambda terms.
In Section~\ref{sec:logrelproof}, we prove the fundamental lemma,
which states that syntactic typing implies semantic typing. Once the
fundamental lemma is established, logical consistency follows as a
trivial corollary.

The downside of a pen and paper proof is that the validity of the
definitions themselves can be dubious at times. When working with a
dependent type theory that has a universe hierarchy, an intuitive
understanding of naive set theory no longer suffices as the size of
sets matters. For example, the partial function representing the
interpretation of types does not immediately correspond to a valid
definition in a proof assistant. We cannot easily define it as a
function as proof assistants such as Coq and Agda only accept total
functions. We cannot naively encode the interpretation function as an
inductive relation either since we violate the strict positivity
constraint. A detailed discussion of how we work around such technical
issues can be distracting. However, what can be equally distracting is
following along a proof where even the validity of the statement is
not trusted. As a result, we clearly mark the ``suspicious''
definitions, and defer the discussion of these definitions
until Section~\ref{sec:logrelmech}, where we give an overview of our
short Coq mechanization proof and the various proof engineering
techniques we adopt to make the mechanization for tractable.

Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
existing literature related to logical consistency about dependent
type theory.

\section{Review of Proof by Logical Relation}
\label{sec:logrelstlc}

\section{Specification of a Dependent Type Theory}
\label{sec:spec}

\section{Semantic Typing}
\label{sec:logreldep}

\section{Consistency}
\label{sec:logreldep}

\section{Mechanization}
\label{sec:logrelmech}

\section{Related Work}
\label{sec:relatedwork}

% Type soundness can be proven through a syntactic
% approach~\citep{syntacticsoundness} as a corollary of two properties:
% progress and preservation. % The syntactic type soundness proof
% % varies in complexity depending on the underlying type
% % system. For example, a type system that tracks information flow would
% % require additional structural rules related to security levels. In
% % this paper, we focus on one specific type of complexity: the
% In Figure~\ref{fig:stlcsoundness}, we summarize the structure of the
% syntactic type soundness proof for the simply typed lambda
% calculus. Each lemma can be proven by structural induction over the
% typing derivation, while using the previous established results as
% lemmas for specific cases that do not immediately follow from the
% induction hypothesis. If we make our language more complex by adding
% full dependent type support, the overall structure remains almost
% identical.

% NbE in Coq
